---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(phyloseq)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggalt)
library(rlang)
library(forcats)
```

# phyloENVS

<!-- badges: start -->
<!-- badges: end -->

The goal of phyloENVS is to provide tailored tools for visualizing and analyzing microbial community data, with a focus on rRNA and mRNA datasets from environmental samples such as soil, ice, and other ecosystems. Built to enhance the functionality of the 'phyloseq' package, it offers customizable and publication-quality visualizations, including abundance plots, diversity metrics, and community composition analyses. The package is designed to streamline workflows for researchers within Environmental Microbiology at Aarhus University.

## Installation

You can install the phyloENVS from [GitHub](https://github.com/) with the devtools package.

``` r
# install.packages("devtools")
devtools::install_github("johanneBiO/phyloENVS")
```

## Example

This is a basic example which shows you can visualize the relative abundance for a microbial community of ice sampled at the Qaanaaq glacier and the Greenland ice sheet:

```{r example}
library(phyloENVS)

data("qaanaaq_rRNA")

qaanaaq_rRNA_sub <- subset_samples(qaanaaq_rRNA,
                                   Transect == "Non-transect") 

vis_abundance(physeq = qaanaaq_rRNA_sub,
              group_x = SampleName,
              group_split = Location,
              level_glom = Phylum,
              level_select = NULL,
              group_select = NULL,
              lower_limit = 2,
              remove_grid = FALSE)
```
