---
title: "Advanced usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{02 Advanced usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how phyloENVS can be used for a larger dataset with replicates.

## Data

We will be using meta-transcriptomic data from the GENEPEASE II project. 

```{r setup, message=FALSE}
library(phyloENVS)
library(phyloseq)
library(dplyr)
library(tidyr)

data("genepease_rRNA")
```

This metadata describes the concentration and time group for each sample. For each combination of concentration and time, we got 3 replicates specified by a group ID as seen for the examples below.

```{r}
head(sample_data(genepease_rRNA), n = 9)
```

## Merge Samples

We want to merge the replicates using `merge_data()` similar to `merge_samples()` from phyloseq. 

```{r}
genepease_rRNA_merged <- merge_data(genepease_rRNA, "Group_ID")
```

The resulting phyloseq object will merge the sample data as well and use the specified column as the new sample names.

```{r}
head(sample_data(genepease_rRNA_merged), n = 9)
```

## Explorative Visualizations

We take a look at the ordination for the merged samples colored by their concentration and shapes by the timepoint.

```{r, message=FALSE, fig.width=7, fig.height=5, out.width="80%", fig.align="center"}
vis_nmds(genepease_rRNA_merged,
         convert_to_rel = TRUE,
         group_color = "Concentration",
         group_shape = "Timepoint")
```

Afterwards, we look at the overall relative abundance.

```{r, fig.width=7, fig.height=5, out.width="80%", fig.align="center"}
vis_abundance(physeq = genepease_rRNA_merged,
              group_x = "Timepoint",
              group_split = "Concentration",
              level_glom = "Phylum",
              lower_limit = 2)
```

From the above, we can see that proteobacteria take over the community for large concentrations. Let's see what happens for the bacteria and the eukaryotes separately. The represented kingdoms can be seen below.

```{r}
tax_table(genepease_rRNA_merged) |>
  as.data.frame() |>
  as_tibble() |>
  select(Kingdom) |>
  pull() |>
  table()
```

We create the table to define a new customize level.

```{r}
kingdom_to_domain <- c(
  "Amorphea" = "Eukaryotes",
  "Archaea" = "Prokaryotes",
  "Archaeplastida" = "Eukaryotes",
  "Bacteria" = "Prokaryotes",
  "Excavata" = "Eukaryotes",
  "Hacrobia" = "Eukaryotes",
  "SAR" = "Eukaryotes"
)
```

We add the new level using `add_level()`.

```{r}
genepease_rRNA_merged <- add_level(physeq = genepease_rRNA_merged, 
                                   level = "Kingdom",
                                   level_name = "Domain",
                                   look_up_table = kingdom_to_domain)
```

Let's look at the relative abundance for eukaryotes on phylum level.

```{r, fig.width=10, fig.height=5, out.width="80%", fig.align="center"}
vis_abundance(physeq = genepease_rRNA_merged,
              group_x = "Timepoint",
              group_split = "Concentration",
              level_glom = "Phylum",
              level_select = "Domain",
              group_select = "Eukaryotes",
              lower_limit = 0.2)
```

In the same manner, we inspect the relative abundance of the prokaryotes, but we will agglomerate on order level to get a better understanding of the taxa taking over for higher concentrations. 

```{r, fig.width=10, fig.height=5, out.width="80%", fig.align="center"}
vis_abundance(physeq = genepease_rRNA_merged,
              group_x = "Timepoint",
              group_split = "Concentration",
              level_glom = "Order",
              level_select = "Domain",
              group_select = "Prokaryotes",
              lower_limit = 3) 
```

## Richness With Specified Measure

Finally, we can visualize the alpha diversity focused on differences between concentrations, but we color by timepoint to look for patterns in the richness with respect to this variable. In the example below, we will use inverse Simpson to measure the diversity. 

```{r, fig.width=7, fig.height=5, out.width="80%", fig.align="center", warning=FALSE}
vis_richness(physeq = genepease_rRNA_merged, 
             group_x = "Concentration", 
             group_color = "Timepoint",
             measures = "InvSimpson")
```

## Univariate Statistical Tests Across Taxa

In the vignette "Introduction to phyloENVS", we show how differences in community structure between groups can be tested with PERMANOVA. It is also possible to make univariate tests for the abundance per taxon. The framework is build into the function `perform_univariate()`. We recommend visiting its documentation. 

```{r, eval=FALSE}
perform_univariate(genepease_rRNA,
                   stats_path = "results/",
                   designs = c("Direction", "Wetness"))
```
