---
title: "Using phyloENVS for Qaanaaq dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(phyloseq)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggalt)
library(rlang)
library(forcats)
```

This vignette demonstrates the core functions of phyloENVS using meta-transcriptomic data from ice samples collected in Qaanaaq, Greenland. The dataset includes samples from both the Qaanaaq Glacier and the Greenland Ice Sheet. In this vignette, we focus only on glacier samples collected along two transects.

As phyloENVS builds upon the phyloseq package for microbial community analysis, we assume that users are familiar with basic phyloseq commands. If you are new to phyloseq, we recommend reviewing [its documentation](https://joey711.github.io/phyloseq/) before proceeding.

## Sampling Design and Environmental Conditions

The transects were established on the Qaanaaq Glacier, with small meltwater rivers on either side. Due to variations in water influence across the glacier surface, the sampling locations along each transect were categorized into three zones:

-   Wet Zone – Closest to the meltwater rivers, experiencing the highest water flow and likely enriched with transported nutrients and microorganisms.
-   Semi-Wet Zone – Intermediate locations, receiving occasional water input but not as saturated as the wet zone.
-   Dry Zone – Farthest from the water sources, with minimal direct meltwater influence, representing more stable glacial ice conditions.

By comparing microbial communities across these zones, we can explore how water availability shapes microbial activity and composition on glacier surfaces.

Let's get started!

```{r setup}
library(phyloENVS)

data("qaanaaq_rRNA")

qaanaaq_rRNA_sub <- subset_samples(qaanaaq_rRNA,
                                   Transect != "Non-transect") 
```

## Ordination

Firstly, we will perform Non-metric Multidimensional Scaling (NMDS) - a statistical method used for visualizing high-dimensional data by reducing it to fewer dimensions (here 2). This ordination will preserve the rank order of Bray-Curtis dissimilarities between samples. By coloring the points by wetness, we can visually investigate whether this environmental factor influence the composition of a community. We will shape the points by transect (1 or 2) to also look for patterns related to each transect.

```{r, message=FALSE, fig.width=7, fig.height=5, out.width="80%", fig.align="center"}
vis_nmds(qaanaaq_rRNA_sub,
         convert_to_rel = TRUE,
         group_color = "Wetness",
         group_shape = "Wetness",
         encircle = TRUE,
         fill_circle = TRUE,
         scale_circle = 0.09,
         scale_plot = 0.3)
```

Be aware, if the read counts are not normalized, we always recommend to normalize to the relative abundance prior to ordiantion (convert_to_rel = TRUE). However, the user might prefer another normalization method or use relative abundances as input, e.g., when filtering out low abundance taxa based on relative abundances.

After generating the NMDS, we plot the Shepard plot to check the goodnees of fit for the ordination method. The Shepard plot compares the original dissimilarities between samples with the dissimilarities between samples in the reduced-dimensional space obtained from NMDS.

```{r, message=FALSE, fig.width=7, fig.height=5, out.width="80%", fig.align="center"}
vis_shepard(qaanaaq_rRNA_sub, 
            convert_to_rel = TRUE)
```

## Abundance

Relative abundance refers to the proportion of each taxon relative to the total number of sequences or observations in a sample, allowing for easy comparison across samples with varying sequencing depths. Again, we will focus on the relative abundances of microbial taxa from our dataset in wet, semi-wet, and dry zones.

```{r, fig.width=7, fig.height=5, out.width="80%", fig.align="center"}
vis_abundance(physeq = qaanaaq_rRNA_sub,
              group_x = "SampleName",
              group_split = "Wetness",
              level_glom = "Phylum",
              level_select = NULL,
              group_select = NULL,
              lower_limit = 2,
              remove_grid = FALSE)
```

For `vis_abundance()`, it is possible to plot the relative abundance for a specific group within a specific level, e.g., only looking at Streptophyta only.  

```{r, fig.width=7, fig.height=5, out.width="80%", fig.align="center"}
vis_abundance(physeq = qaanaaq_rRNA_sub,
              group_x = "SampleName",
              group_split = "Wetness",
              level_glom = "Phylum",
              level_select = "Phylum",
              group_select = "Streptophyta",
              lower_limit = 0.5,
              remove_grid = FALSE)
```

When looking into the relative abundance for a specific phyla, it might be useful to change the level to agglomerate counts, e.g., on order level.

```{r, fig.width=7, fig.height=5, out.width="80%", fig.align="center"}
vis_abundance(physeq = qaanaaq_rRNA_sub,
              group_x = "SampleName",
              group_split = "Wetness",
              level_glom = "Order",
              level_select = "Phylum",
              group_select = "Chytridiomycota",
              lower_limit = 0,
              remove_grid = FALSE)
```

We can also group taxa into specified groups with `add_level()`. For example, we might want to add a new taxonomical level for the different kingdoms in order to inspect the relative abundances of eukaryotes and prokaryotes seperately. First, we should define a table based on the kingdoms represented in the data. 

```{r}
tax_table(qaanaaq_rRNA_sub) |>
  as.data.frame() |>
  as_tibble() |>
  select(Kingdom) |>
  pull() |>
  table()
```

We create the table.

```{r}
kingdom_to_domain <- c(
  "Amorphea" = "Eukaryotes",
  "Archaeplastida" = "Eukaryotes",
  "Bacteria" = "Prokaryotes",
  "Hacrobia" = "Eukaryotes",
  "SAR" = "Eukaryotes"
)
```

We add the domain level.

```{r}
qaanaaq_rRNA_sub <- add_level(physeq =qaanaaq_rRNA_sub, 
                              level = "Kingdom", 
                              level_name = "Domain", 
                              look_up_table = kingdom_to_domain)
```

Let's look at the relative abundance for prokaryotes on phylum level.

```{r, fig.width=7, fig.height=5, out.width="80%", fig.align="center"}
vis_abundance(physeq = qaanaaq_rRNA_sub,
              group_x = "SampleName",
              group_split = "Wetness",
              level_glom = "Phylum",
              level_select = "Domain",
              group_select = "Prokaryotes",
              lower_limit = 0.5,
              remove_grid = FALSE)
```

## Richness

By plotting alpha diversity through richness, you can observe how many distinct taxa are present in each sample, which is crucial for understanding community structure and the complexity of ecosystems.

```{r, fig.width=7, fig.height=5, out.width="80%", fig.align="center"}
vis_richness(physeq = qaanaaq_rRNA_sub, 
             group_x = "Wetness", 
             group_color = "Wetness")
```
